---
title: "PA1_template"
author: "Trey Storer"
date: "December 15, 2015"
output: html_document
---
#About
This R Markdown document will complete the work necessary for submission of Assignment 1 of Reproducible Research measuring the results of step tracking devices in five minute intervals over two days.

```{r standards, echo=TRUE}

echo=TRUE
options(scipen = 1)
```


##Loading and Preprocessing the data:
Connect to the data, download, unzip and read the raw data


```{r data processing, echo=TRUE}

setwd("C:/Users/sc58263/Documents/Coursera/ReproducibleResearch")
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
  temp <- tempfile()
  download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
  unzip(temp)
  unlink(temp)
}

activity <- read.csv("activity.csv")

str(activity)
```

Based on the results of the data structure, it is clear the dates were not formatted correctly. So, we will correct that:

```{r, date correction, echo=TRUE}
activity$date<- as.Date(activity$date)
str(activity)
rmactivity<- activity[which(!is.na(activity$steps)),]
```

##Calculate and plot the histogram for the steps taken per day.

```{r calculate daily steps, echo=TRUE}
daily <- aggregate(steps ~ date, activity, sum)
hist(daily$steps, col= "lightgreen", main = paste("Total Steps Each Day", ylab = "Number of Steps"))
```

```{r calculate mean and median, echo=TRUE}
stepmean<- mean(daily$steps)
stepmedian<- median(daily$steps)
```

The mean is `r stepmean` and the median is`r stepmedian`.

##What is the average daily activity pattern?
* Calculate the average steps per interval
* Plot the average steps per interval
* Find the interval with the highest number of steps

```{r generate the average steps by interval, echo=TRUE}
intervalsteps <- aggregate(steps ~ interval, rmactivity, mean)
plot(intervalsteps$interval,intervalsteps$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")
max_interval <- intervalsteps[which.max(intervalsteps$steps),1]
max_steps <- intervalsteps[which.max(intervalsteps$steps),2]

```

The maximum average steps in any interval is `r max_steps`, occurring at `r max_interval`.

##Imputing Missing Values
There are many records with missing values, so we need to be sure that the reporting is not adversely affected by these gaps.
As a reminder, let's check the number of missing values in the steps columns.
```{r calculate number of missing values, echo=TRUE}
sum(is.na(activity$steps))
```
Further, we will check the missing values overall to ensure we have found all of the missing values.
```{r calculate missing values for the entire table, echo=TRUE}
sum(is.na(activity))
```

Now, we will create a strategy to impute the number of steps for each interval with missing values.  We will use the mean of the steps for matching intervals across the days with valid measurements.

```{r impute a value for missing entries, echo=TRUE}
imputed_data <- transform(activity, steps = ifelse(is.na(activity$steps), intervalsteps$steps[match(activity$interval, intervalsteps$interval)], activity$steps))
sum(is.na(imputed_data))
```

There are no longer any missing values.

Generate a new Histogram based on the data with imputed values in place of the missing values.

```{r calculate imputed daily steps, echo=TRUE}
imputed_daily <- aggregate(steps ~ date, imputed_data, sum)
hist(imputed_daily$steps, col= "lightblue", main = paste("Total Steps Each Day", lab = "Number of Imputed Steps"))
```

We should compare the old values with the new Imputed values to see how our effort changed the results.

Calculate the new mean and median:
```{r calculated the imputed mean and median, echo=TRUE}
imputed_mean<- mean(imputed_daily$steps)
imputed_median<- median(imputed_daily$steps)
```

The new, imputed mean is `r imputed_mean`, and the new imputed medians is `r imputed_median`.

```{r calculate variance old to new}
mean_var<- imputed_mean - stepmean
med_var<- abs(imputed_median - stepmedian)
```

The impact was minimal.  The variance between the old and new mean was `r mean_var` steps and the variance between the medians was `r med_var` steps.

## How does activity change based on the day of the week, particularly between weekdays and weekends?

We will assign our records to either weekday or weekend based on the date the information was recorded, then summarize the results by day of the week and interval.

```{r day of week comparison, echo=TRUE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
 
```

It appears that activity starts earlier on weekdays, but stays at a higher leverl throughout the day on weekends.

